---
title: 'Chapter 4: Categories and Curves'
author: "Paulo Magalang"
date: "2023-10-17"
output: html_document
---

Lecture: https://www.youtube.com/watch?v=F0N4b7K_iYQ&list=PLDcUM9US4XdPz-KxHM4XHt7uUVGWWVSus&index=5

```{r, message = FALSE, warning = FALSE}
library(rethinking)
library(tidyverse)
```


Need to incorporate causal thinking into:

1. How we draw statistical models

* generative model + multiple estimands -> multiple estimators

2. How we process the results

* only very simple causal estimates show up in summary tables


Categorical variables and Curves to build causal estimators

* categories: indicator and index variables

* splines and other structures 


Categories

* discrete, unordered types

* stratify by category: fit a separate line for each


* how are height (H), weight (W), and sex (S) causally related and statistically
related?

* consider the intervention to determine causal relationship in DAG

S -> H, W

H -> W

* influence on weight by height is direct but influence on weight by sex is
both direct and indirect

$$
H = f_H(S) \\
W = f_W(H, S)
$$

```{r}
set.seed(1)

# S=1 female; S=2 male
sim_HW <- function(S, b, a) {
  N <- length(S)
  H <- ifelse(S==1, 150, 160) + rnorm(N, 0, 5)
  W <- a[S] + b[S]*H + rnorm(N, 0, 5) # for each S, different height-weight line
  data.frame(S, H, W)
}

S <- rbern(100) + 1
dat <- sim_HW(S, b = c(0.5, 0.6), a = c(0, 0))
head(dat)
```

Different causal questions need different statistical models:

1. Causal effect of H on W?

2. Causal effect of S on W? (direct and indirect effects)

3. Direct causal effect of S on W? (direct effect only)

* stratify by S: different estimate for each value S can take

Categorial OWLs

Several ways to code categorical variables: indicator (0/1) var or index var (1, 2, 3, ...)

Example:

$$
\alpha = [\alpha_1, \alpha_2, \alpha_3, \alpha_4] \\
y_i \sim Norm(\mu_i, \sigma) \\
\mu_i = \alpha_{COLOR}[i]
$$

Using Index Variables

$$
W_i \sim Norm(\mu_i, \sigma) \\
\mu_i = \alpha_{S[i]} \\
\alpha_j \sim Norm(60, 10) \\
\alpha = [\alpha_1, \alpha_2] \\
\sigma \sim Uni(0,10)
$$

where $S[i]$ is the sex of the i-th person in data frame


Testing

What is the total causal effect of sex? Causal effect of sex is differece made intervening:

```{r}
set.seed(1)

# female sample
S <- rep(1, 100)
simF <- sim_HW(S, b = c(0.5, 0.6), a = c(0, 0))

# male sample
S <- rep(2, 100)
simM <- sim_HW(S, b = c(0.5, 0.6), a = c(0, 0))

# effect of sex (male - female)
mean(simM$W - simF$W)
```

Now run the estimating model and synthetic sample

```{r}
set.seed(1)

# observe sample
S <- rbern(100) + 1
dat <- sim_HW(S, b = c(0.5, 0.6), a = c(0, 0))

# estimate prior
m_SW <- quap(
  alist(
    W ~ dnorm(mu, sigma),
    mu <- a[S],
    a[S] ~ dnorm(60, 10),
    sigma ~ dunif(0, 10)
  ), data = dat)
precis(m_SW, depth = 2)
```


Analyze the real sample

```{r}
data(Howell1)
d <- Howell1
d <- d[d$age >= 18, ]
dat <- list(
  W = d$weight,
  S = d$male + 1) # S=1 female, S=2 male

m_SW <- quap(
  alist(
    W ~ dnorm(mu, sigma),
    mu <- a[S],
    a[S] ~ dnorm(60, 10),
    sigma ~ dunif(0, 10)
  ), data = dat)
```

```{r}
# posterior mean W
post <- extract.samples(m_SW)

df <- data.frame(c(post$a[, 1], post$a[, 2]),
                 c(rep("W", length(post$a[, 1])), rep("M", length(post$a[, 2])))) 
colnames(df) <- c("mean_W", "Sex")

df %>% ggplot(aes(x = mean_W, fill = Sex)) + geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("blue", "red")) +
  xlab("posterior mean weight (kg)")
  

#dens(post$a[, 1], xlim = c(39, 50), lwd = 3, col = 2,
#     xlab = "posterior mean weight (kg)")
#dens(post$a[, 2], lwd = 3, col = 4, add = TRUE)
```

```{r}
# posterior W distributions
W1 <- rnorm(1000, post$a[, 1], post$sigma)
W2 <- rnorm(1000, post$a[, 2], post$sigma)

df <- data.frame(c(W1, W2),
                 c(rep("W", length(W1)), rep("M", length(W2)))) 
colnames(df) <- c("Weight", "Sex")

df %>% ggplot(aes(x = Weight, fill = Sex)) + geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("blue", "red")) +
  xlab("posterior predicted weight (kg)")

#dens(W1, xlim = c(20, 70), ylim = c(0, 0.085), lwd = 3, col = 2)
#dens(W2, lwd = 3, col = 4, add = TRUE)
```

* need to compute contrast (difference between the categories), never legitimate
to compare overlap in paramters or lines

```{r}
# causal contrast (in means)
mu_contrast <- post$a[, 2]- post$a[, 1]

dens(mu_contrast, xlim = c(3, 10), lwd = 3, col = 1,
     xlab = "posterior mean weight contrast (kg)")
```

```{r}
set.seed(1)

# posterior @ distributions
W1 <- rnorm(1000, post$a[, 1], post$sigma)
W2 <- rnorm(1000, post$a[, 2], post$sigma)

# contrast
W_contrast <- W2 - W1
dens(W_contrast, xlim = c(-25, 35), lwd = 3, col = 1,
     xlab = "posterior weight contrast (kg)")

# proportion above zero
sum(W_contrast > 0) / 1000
# proportion below zero
sum(W_contrast < 0) / 1000
```

Direct causal effect of S on W? Need to "block" association through H. partial out

```{r}
set.seed(1)
S <- rbern(100) + 1
dat <- sim_HW(S, b = c(0.5, 0.5), a = c(0, 10))
# indirect: H influences W differently (slopes)
# direct: W differs net any slope differences
```

$$
W_i \sim Norm(\mu_i, \sigma) \\
\mu_i = \alpha_{S[i]} + \beta_{S[i]}(H_i - \bar{H}) \\
\alpha = [\alpha_1, \alpha_2] \\
\beta = [\beta_1, \beta_2]
$$

$H_i - \bar{H}$ is the centered height so that $\alpha$ is the average weight
of a person with average height, makes it easier to define priors for $\alpha$ 

```{r}
# analyze the sample
data("Howell1")
d <- Howell1
d <- d[d$age >= 18, ]
dat <- list(
  W = d$weight,
  H = d$height,
  Hbar = mean(d$height),
  S = d$male + 1) # S=1 female, S=2 male

m_SHW <- quap(
  alist(
    W ~ dnorm(mu, sigma),
    mu <- a[S] + b[S]*(H - Hbar),
    a[S] ~ dnorm(60, 10),
    b[S] ~ dunif(0, 1),
    sigma ~ dunif(0, 10)
  ), data = dat)
```

```{r}
xseq <- seq(from = 130, to = 190, len = 50)

muF <- link(m_SHW, data = list(S = rep(1, 50), H = xseq, Hbar = mean(d$height)))
#lines(xseq, apply(muF, 2, mean), lwd = 3, col = 2)

muM <- link(m_SHW, data = list(S = rep(2, 50), H = xseq, Hbar = mean(d$height)))
#lines(xseq, apply(muM, 2, mean), lwd = 3, col = 4)

mu_contrast <- muF - muM
plot(NULL, xlim = range(xseq), ylim = c(-6, 8), xlab = "height (cm)",
     ylab = "weight contrast (F-M)")
for (p in c(0.5, 0.6, 0.7, 0.8, 0.9, 0.99))
  shade(apply(mu_contrast, 2, PI, prob = p), xseq)
abline(h = 0, lty = 2)
```

nearly all of the causal effect of S acts through H


Most relationships are nonlinear, can easily fit linear models but not inherently
mechanistic

Two popular strategies

1. polynomial linear models

$\mu_i = \alpha + \beta_1 x_i + \beta_2 x_i^2$

explosive uncertainty at edges, strange symmetry

no local smoothing, only global, technically linear

2. splines and generalized additive models -- less awful

basis-splines: wiggly function built from many local functions

$ \mu_i = \alpha + w_1 B_{i,1} + w_2 B_{i, 2} + ...$

sum up basis functions to form spline, basis function governed by respective weight

local functions make splines better than polynomials, but equally geocentric

adding scientific data would help


"Full Luxury Bayes"

use one model of entire causal system instead of building two separate models

use joint posterior to compute each estimand

```{r}
m_SHW_full <- quap(
  alist(
    # weight
    W ~ dnorm(mu, sigma),
    mu <- a[S] + b[S]*(H-Hbar),
    a[S] ~ dnorm(60, 10),
    b[S] ~ dunif(0, 1),
    sigma ~ dunif(0, 10),
    
    # height
    H ~ dnorm(nu, tau),
    nu <- h[S],
    h[S] ~ dnorm(160, 10),
    tau ~ dunif(0, 10)
  ), data = dat)
```

```{r}
post <- extract.samples(m_SHW_full)
Hbar <- dat$Hbar
n <- 1e4

with(post, {
  # simulate W for S=1
  H_S1 <- rnorm(n, h[, 1], tau)
  W_S1 <- rnorm(n, a[, 1] + b[, 1]*(H_S1 - Hbar), sigma)
  
  # simulate W for S=2
  H_S2 <- rnorm(n, h[, 2], tau)
  W_S2 <- rnorm(n, a[, 2] + b[, 2]*(H_S2 - Hbar), sigma)
  
  # compute contrast
  W_do_S <<- W_S2 - W_S1
})
```

$p(W | do(S))$ where do() represents an intervention

```{r}
# automated way
#HWsim <- sim(m_SHW_full,
#             data = list(S = c(1, 2)),
#             vars = c("H", "W"))
#W_do_S_auto <- HWsim$W[, 2] - HWsim$W[, 1]
```














