---
title: "Week 2 Problem Set"
author: "Paulo Magalang"
date: "2023-10-11"
output: html_document
---

```{r, eval = FALSE, warning= FALSE}
library(rethinking)
library(tidyverse)
library(ggplot2)
```

4E1. In the model definition below, which line is the likelihood?

$$
y_i \sim Normal(\mu, \sigma) \\
\mu \sim Normal(0, 10) \\
\sigma \sim Exponential(1)
$$

The first line is the likelihood.

4E2. In the model definition above, how many parameters are in the posterior distribtution?

There are two paramters but $y_i$ can be observable or unobservable which can contribute
to the number of parameters.

4E3. Using the model definition above, write down the appropriate form of Bayes'
theorem that includes the proper likelihood and priors.

$$
P(\mu, \sigma | y_i) = \frac{P(y_i | \mu, \sigma)P(\mu, \sigma)}{P(y_i)}
$$

4E4. In the model definition below, which line is the linear model?

$$
y_i \sim Normal(\mu, \sigma) \\
\mu_i = \alpha + \beta x_i \\
\alpha \sim Normal(0, 10) \\
\beta \sim Normal(0, 1) \\
\sigma \sim Exponential(2)
$$

$\mu_i = \alpha + \beta x_i$ is the linear model.

4E5. In the model definition above, how many parameters are in the posterior distribution?

The posterior can be written as $P(\alpha, \beta, \sigma | y_i, \mu_i)$


4M1. For the model defintion below, simulate observed y values from the prior (not the
posterior).

$$
y_i \sim N(\mu, \sigma) \\
\mu \sim N(0, 10) \\
\sigma \sim Exp(1)
$$

```{r}
set.seed(12345)

n <- 1e3
mu <- rnorm(n, 0, 10)
sigma <- rexp(n, 1)
```

```{r}
y_sample <- rnorm(n, mu, sigma)
data.frame(y_sample) %>% ggplot(aes(x = y_sample)) + geom_histogram(binwidth = 1) +
  xlab("Simulated y values")
```

4M2. Translate the model just above into a quap formula.

```{}
q_4M2 <- quap(
  alist(
    y ~ dnorm(mu, sigma),
    mu ~ dnorm(0, 10),
    sigma ~ dexp(1),
  ), data = df
)
```

4M3. Translate the quap model formula below into a mathematical model definition.

```{}
y ~ dnorm(mu, sigma),
mu <- a + b*x,
a ~ dnorm(0, 10),
b ~ dunif(0, 1),
sigma ~ dexp(1)
```

$$
y_i \sim N(\mu_i, \sigma) \\
\mu_i = a + bx_i \\
a \sim N(0, 10) \\
b \sim Uniform(0, 1) \\
\sigma \sim Exp(1)
$$

4M4. A sample of students is measured for height each year for 3 years. After the third year,
you want to fit a linear regression predicting height using year as a predictor. Write
down the mathematical model definition for this regression, using any variable names
and priors you choose. Be prepared to defend your choice of priors.

$$
h_i \sim Norm(\mu_i, \sigma) \\
\mu_i = \alpha + \beta (t_i - t_{birth}) \\
\alpha \sim  Uni(0, 100)\\
\beta \sim Uni(0, 10)\\
\sigma \sim Exp(1)
$$


4M5. Now suppose I tell you that the variance among heights for the students of the same
age is never more than 64cm. How does this lead you to revise your priors?

4M7. Refit model `m4.3` from the chapter, but omit the mean weight `xbar` this time. 
Compare the new model's posterior to that of the original model. In particular, look at
the covariance among tha parameters. What is different? Then compare the posterior predictions
of both models.

4H1. The weights listed below were recorded in the !Kung census, but heights were not recorded
for these individuals. Provide predicted heights and 89% intervals for each of these individuals.
That is, fill in the table below, using model-based predictions.




















